Perceived value of video games, but not hours played, predicts mental well-being in casual adult Nintendo players

### 2024-12-16 ms.qmd
R1 update
Merge branch 'main' of github.com:digital-wellbeing/noa-pilot-ms

# Conflicts:
#	README.md

---
title: "Perceived value of video games, but not hours played, predicts mental well-being in causal adult Nintendo players"
shorttitle: "Nintendo gaming and well-being"
author:
  - name: Nick Ballou
    corresponding: false
    orcid: 0000-0003-4126-0696
    email: nick@nickballou.com
    affiliations:
      - name: University of Oxford
        department: Oxford Internet Institute
    # roles:
    #   - conceptualization
    #   - data curation
    #   - methodology
    #   - formal analysis
    #   - writing

@@

  - name: Andrew K Przybylski
    corresponding: true
    email: andy.pryzbylski@oii.ox.ac.uk
    orcid: 0000-0003-4126-0696
    affiliations:
      - name: University of Oxford
        department: Oxford Internet Institute
    # roles:
    #   - conceptualization
    #   - funding acquisition
    #   - project administration
    #   - editing
abstract: "Studies on video games and well-being often rely on self-report measures or data from a single game. Here, we study how 703 casually-engaged US adults’ time spent playing for over 140,000 hours across 150 Nintendo Switch games relates to their life satisfaction, affect, depressive symptoms, and general mental well-being. We replicate previous findings that playtime over the past two weeks does not predict well-being, and extend these findings to a wider range of timescales (one hour to one year). Equivalence tests were inconclusive, and thus we do not find evidence of absence, but results suggest that relationships, if present and practically meaningful, likely dissipate within two hours of gameplay. Our non-causal findings suggest substantial confounding would be needed to shift a meaningful true effect to the observed null. Although playtime was not related to well-being, players’ assessments of the value of game time—so called gaming life fit—was. Results emphasise the importance of defining the gaming population of interest, collecting data from more than one game, and focusing on how players integrate gaming into their lives rather than the amount of time spent."
keywords: [video games, digital trace data, well-being]
author-note:
  # Disclosures condensed to one paragraph, but you can start a field with two line breaks to break them up: \n\nNew Paragraph
  disclosures:
    study-registration: This study was registered on the Open Science Framework (https://osf.io/sjqyt).
    # Acknowledge and cite data/materials to be shared.
    data-sharing: Data and materials are available at https://osf.io/6xkdg/. 
    # Example: [Author name] has been a paid consultant for Corporation X, which funded this study.
    conflict-of-interest: Data for this study was provided by Nintendo of America. Nintendo of America had no role in the design of the study, the analysis of the data, or decision to publish any results. The authors declare no other competing interests. 
    # Example: This study was supported by Grant [Grant Number] from [Funding Source].
    financial-support: This research was supported by Huo Family Foundation and the UK Economic and Social Research Council (ES/W012626/1). KM was supported by Forskningsrå det för hälsa, arbetsliv och välfärd (2021-01284).
    # Example: The authors are grateful to [Person] for [Reason].
    gratitude: ~
    # Example. Because the authors are equal contributors, order of authorship was determined by a fair coin toss.
    authorship-agreements: ~
bibliography: references.bib
floatsintext: true # If true, tables and figures are mingled with the text instead of listed at the end of the document.
numbered-lines: true # Numbered lines (.pdf and .docx only)
meta-analysis: false
suppress-credit-statement: true
format:
  apaquarto-docx: default
  # apaquarto-html:
  #   documentmode: man
  #   code-fold: true
  #   code-summary: "Show the code"
  #   wordcount-html: default
  apaquarto-pdf:
    documentmode: jou
---

@@

Video games played on smartphones, computers, or home consoles are now a widely pursued form of leisure that involves social interaction, creativity, problem-solving, and growth [@BourgonjonEtAl2016Players]. Major firms like Nintendo have sold hundreds of millions of games consoles in recent years [@NintendoJapan2024Top] and online platforms such as Steam regularly attract upwards of 30 million players online at any given time [@SteamCharts2024SteamCharts]. This staggering investment of human attention and behaviour has inspired both national [@AmericanPsychiatricAssociation2013diagnostic] and international [@WorldHealthOrganization2018international] health bodies to focus on play as potential contributor to psychopathology.

@@

An important frontier for the field, therefore, is to collect holistic digital trace data that reflects behavior in not just one game, but all games played (which may include various games on one platform, such as Nintendo Switch or Steam, or all games played across several platforms a player uses). At present, our understanding of even basic phenomena such as the true volume of play in different demographics rely on the same inaccurate self-report data, itself often provided by market research firms using opaque methodologies. Capturing play at the platform level represents one step towards this goal of studying a player's complete gaming diet. To our knowledge, the only study to investigate player health with platform-level digital trace data is [@BallouEtAl2024Registered], which found no meaningful relationships between Xbox playtime and well-being over three months. The current study complements this earlier work by investigating how well-being relates to play on an alternative platform, Nintendo Switch, which has a more causal audience and a distinct library of games, and by analyzing a longer window of play behavior spanning up to one year.

@@

Our participant flow is shown in @fig-sankey. We recruited participants from Prolific, a participant recruitment platform, who were: (1) at least 18 years old, (2) proficient in English, (3) residents of the US, and (4) active video game players (self-defined, based on a \>0 response to Prolific's built-in screening item "How many hours per week do you play video games on average?"). We first distributed a screening questionnaire to 7649 participants asking which video game platforms they were active on; of these, 4184 indicated that they played games on Nintendo Switch. Participants who played on Nintendo Switch were active on 1.8 additional platforms on average (e.g., also using Playstation and iOS); their Nintendo playtime therefore captures only an (unknown) portion of their overall gaming activity. We return to this limitation in the Discussion.

@@

Participants completed a self-report survey that took on average 22 minutes to complete. The survey included background factors such as demographics and life circumstances, a series of cognitive tasks, as well as self-report measures of time use, and motivations for video game play. In this paper, we focus on the playtime data and the self-reported well-being measures, and detail only those measures below. Other measures were intended for exploratory analyses beyond the scope of the current paper. All study data and materials are available on the OSF (see supplementary materials).

@@

```{r}
#| label: fig-playtime
#| fig-cap: "Description of playtime in the sample. Panel A shows the distribution of playtime across players for each observation period, with the mean shown in red; observation periods below the dashed line are shown as total hours, whereas periods above the dashed line reflect hours of player per day. Panel B shows the proportion of players who logged at least one play session in that period. Raw descriptive statistics are available in the supplementary materials (tables-figures/playtimeDescriptives.csv)."
#| fig-width: 8
#| fig-height: 4
#| fig-align: center
#| fig-pos: "t"
#| apa-twocolumn: true

@@

However, we did find evidence of direct positive relationships linking life fit to well-being separate from playtime (@fig-lifefit): Those who believe gaming to be beneficial to their lives tend to are also more likely to report higher levels of well-being, regardless of how much they play. Across 48 models, we observed relationships between well-being and a 1 point change in life fit ranging from `r min(directEsts$estimate)` to `r max(directEsts$estimate)` (median = `r round(median(directEsts$estimate), 3)`; all ps \< .001). In sensitivity analyses, life fit was both a significant predictor of daily life need satisfaction (`r nsMod$pretty`) and frustration (`r nfMod$pretty`), and largely remained a significant predictor of well-being when controlling for need satisfaction and frustration (median b = `r round(median(fittedNSModels$estimate), 3)`, median p value = `r round(median(fittedNSModels$p.value), 3)`). This suggests that gaming life fit may account for well-being above and beyond someone's level of fulfillment in life as a whole.

@@

Although we did not intend our study to test the causal question most critical to heated debates surrounding video game engagement and global health, our study is a concrete step towards long-repeated calls for greater transparency and robustness in games research, having independently recruited a large sample of video game players [as opposed relying on recruitment through games companies themselves; e.g. @JohannesEtAl2021video], collected validated measures of well-being, joined these with digital trace data and made minimal adjustments for age, gender, employment, and education. Using these methods, we did not uncover robust or consistent relationships between time spent playing and various mental well-being outcomes.

@@

To further elucidate this point, we conducted brief simulation tests to ascertain how strong such confounding might need to be (see Supplementary Materials). For example, if the true standardized effect of playtime on mental health was a moderate .2 SDs per additional hour of daily playtime, a confound C would need to be a very strong cause of both X ($beta = .5$) and Y ($beta = -.5$ ) to bias the true .2 effect to null. While we do not claim this is impossible, we do believe it unlikely. Approaching the topic along these lines---identifying confounds, testing the presence or absence of correlations for their sensitivity to potential confounds, and systematically identifying factors that do (not) cause playtime and well-being---can help us achieve more systematic progress [@BallouEtAl2024How].This work can be bolstered by qualitative research aimed at more fully mapping the causal system and by substantive theory development with greater specificity in the aspects of media use expected to produce effects, the hypothesized causal relationships, boundary conditions, and so forth [@MagnussonEtAl2024Harmful; @Ballou2023Manifesto; @Coenen2023Lost; @EronenBringmann2021theory].

@@

As the field progresses, however, differences in the level of engagement pose major challenges for study sampling and generalisability. Calls for representative samples need to specify the population of reference: should this be the general population (of whom many do not play games), people who play any games at all (of whom many do not play the games for which researchers have data access), people who play the particular game or platform of interest (of whom many may be only minimally engaged), or something else entirely? In the field's quest for more generalizable results, this will be a critical issue.

@@

While we are quick to caution that this is a preliminary finding that should not be relied upon without further validation, our data provide some initial evidence that effects of raw playtime might materialize and fade within a few hours---point estimates indicated that playtime was more strongly linked with greater well-being in the 1-2 hours prior to survey completion. This finding is compatible with various causal explanations: for example, players who recently played are more likely be in a period of leisure time, which would be expected to generate more positive feelings than in peers doing obligatory activities such as work. However, if researchers do expect to see positive effects of gaming, our data suggest that they may need to search for highly proximal effects directly during and following a play session [e.g., @VuorreEtAl2024Affective].

Should this finding be upheld, it would go a long way towards explaining the juxtaposition of previous null findings from studies that related well-being to playtime over timescales such as two weeks [@VuorreEtAl2022Time], one month [@SibillaEtAl2021Harmonious], six months [@WeinsteinEtAl2017prospective], and one year [@KowertEtAl2015Psychosocial]---alongside other studies showing small relationships over the course of a day [@Allen2020gaming; @Razum2023Daily]. For most players, it may be the case that gaming is a recovery activity that helps to manage day-to-day stresses and mood fluctuations, without necessarily having substantial long-term impacts. The majority of players have several options for activities in their environment that would have comparable effects on their well-being. Such activities are thus “exchangeable”, serving the same short-term goals without consequences for long-term adjustment. Studying relationships over the course of hours has to date been possible largely only in laboratory settings—rarely have researchers had access to session-level data of naturalistic behavior that they could link to momentary well-being.

@@

There are four limitations and constraints on generalisability the merit mention. First, we could not analyse digital trace data generated when players engaged with third-party titles (i.e. games not published by Nintendo or its closely associated companies). As a result, our findings only hold for similar Nintendo games (75% of which were rated for everyone or everyone over 10), and it is not possible to rule out the possibility that third-party games with different content or themes might show a different pattern of effects.

@@

The idea that time spent playing is the key ingredient in how games impact well-being will be with us for some time. Although our study was designed to test an association rather than a causal link, it challenges the notion that simply playing more affects well-being, for better or for worse. The correlations we observed were mostly too small to practically matter. Moreover, we show that profound confounding would be required to suppress a true causal effect to account for the null associations we report. This is improbable but not impossible, and we believe our results lend weight to calls for scholars and health practitioners to embrace the gradual shift towards focusing on the quality, rather than quantity of video game play as the key factor for player health---in the first instance, this might be as simple as asking players how they feel gaming fits into and differentially affects other aspects of their life such as work/school, sleep, emotion regulation, and so on. This can be done in the context of self-report questionnaires, qualitative research, diagnostic interviews, and conversations between players and loved ones. To parlay this feedback from players into more systematic theory development, we encourage researchers to familiarize themselves with the breadth of proposed mechanisms through which different patterns—rather than hours—of gaming can affect people [e.g., @BallouEtAl2024How].

>>>>

---
title: "Perceived value of video games, but not hours played, predicts mental well-being in casual adult Nintendo players"
shorttitle: "Nintendo gaming and well-being"
author:
  - name: Nick Ballou
    corresponding: false
    orcid: 0000-0003-4126-0696
    email: nick@nickballou.com
    affiliations:
      - name: University of Oxford
        department: Oxford Internet Institute
    # roles:
    #   - conceptualization
    #   - data curation
    #   - methodology
    #   - formal analysis
    #   - writing

@@

  - name: Andrew K Przybylski
    corresponding: true
    email: andy.pryzbylski@oii.ox.ac.uk
    orcid: 0000-0003-4126-0696
    affiliations:
      - name: University of Oxford
        department: Oxford Internet Institute
    # roles:
    #   - conceptualization
    #   - funding acquisition
    #   - project administration
    #   - editing
abstract: "Studies on video games and well-being often rely on self-report measures or data from a single game. Here, we study how 703 casually-engaged US adults’ time spent playing for over 140,000 hours across 150 Nintendo Switch games relates to their life satisfaction, affect, depressive symptoms, and general mental well-being. We replicate previous findings that playtime over the past two weeks does not predict well-being, and extend these findings to a wider range of timescales (one hour to one year). Equivalence tests were inconclusive, and thus we do not find evidence of absence, but results suggest that practically meaningful effects lasting more than two hours after gameplay are unlikely. Our non-causal findings suggest substantial confounding would be needed to shift a meaningful true effect to the observed null. Although playtime was not related to well-being, players’ assessments of the value of game time—so called gaming life fit—was. Results emphasise the importance of defining the gaming population of interest, collecting data from more than one game, and focusing on how players integrate gaming into their lives rather than the amount of time spent."
keywords: [video games, digital trace data, well-being]
author-note:
  # Disclosures condensed to one paragraph, but you can start a field with two line breaks to break them up: \n\nNew Paragraph
  disclosures:
    study-registration: This study was registered on the Open Science Framework (https://osf.io/sjqyt).
    # Acknowledge and cite data/materials to be shared.
    data-sharing: Data and materials are available at https://osf.io/6xkdg/. 
    # Example: [Author name] has been a paid consultant for Corporation X, which funded this study.
    conflict-of-interest: Data for this study was provided by Nintendo of America. Nintendo of America had no role in the design of the study, the analysis of the data, or decision to publish any results. The authors declare no other competing interests. 
    # Example: This study was supported by Grant [Grant Number] from [Funding Source].
    financial-support: This research was supported by Huo Family Foundation and the UK Economic and Social Research Council (ES/W012626/1). KM was supported by Forskningsrå det för hälsa, arbetsliv och välfärd (2021-01284).
    # Example: The authors are grateful to [Person] for [Reason].
    gratitude: ~
    # Example. Because the authors are equal contributors, order of authorship was determined by a fair coin toss.
    authorship-agreements: ~
bibliography: references.bib
floatsintext: true # If true, tables and figures are mingled with the text instead of listed at the end of the document.
numbered-lines: true # Numbered lines (.pdf and .docx only)
meta-analysis: false
suppress-credit-statement: true
format:
  apaquarto-docx: default
  # apaquarto-html:
  #   documentmode: man
  #   code-fold: true
  #   code-summary: "Show the code"
  #   wordcount-html: default
  apaquarto-pdf:
    documentmode: jou
---

@@

Video games played on smartphones, computers, or home consoles are now a widely pursued form of leisure that involves social interaction, creativity, problem-solving, and growth [@BourgonjonEtAl2016Players]. Major firms like Nintendo have sold hundreds of millions of games consoles in recent years [@NintendoJapan2024Top] and online platforms such as Steam regularly attract upwards of 30 million players online at any given time [@SteamCharts2024SteamCharts]. This staggering investment of human attention and behavior has inspired both national [@AmericanPsychiatricAssociation2013diagnostic] and international [@WorldHealthOrganization2018international] health bodies to focus on play as potential contributor to psychopathology.

@@

An important frontier for the field, therefore, is to collect holistic digital trace data that reflects behavior in not just one game, but all games played (which may include various games on one platform, such as Nintendo Switch or Steam, or all games played across several platforms a player uses). At present, our understanding of even basic phenomena such as the true volume of play in different demographics rely on the same inaccurate self-report data, itself often provided by market research firms using opaque methodologies. Capturing play at the platform level represents one step towards this goal of studying a player's complete gaming diet. To our knowledge, the only study to investigate player health with platform-level digital trace data is [@BallouEtAl2024Registered], which found no meaningful relationships between Xbox playtime and well-being over three months. The current study complements this earlier work by investigating how well-being relates to play on an alternative platform, Nintendo Switch, which has a more casual audience and a distinct library of games, and by analyzing a longer window of play behavior spanning up to one year.

@@

Our participant flow is shown in @fig-sankey. We recruited participants from Prolific, a participant recruitment platform, who were: (1) at least 18 years old, (2) proficient in English, (3) residents of the US, and (4) active video game players (self-defined, based on a \>0 response to Prolific's built-in screening item "How many hours per week do you play video games on average?"). We first distributed a screening questionnaire to 7649 participants asking which video game platforms they were active on; of these, 4184 indicated that they played games on Nintendo Switch. Participants who played on Nintendo Switch were active on 1.8 additional platforms on average (e.g., also using Playstation and iOS); Nintendo playtime therefore captures only an unknown portion of overall gaming activity. We return to this limitation in the Discussion.

@@

Participants completed a self-report survey that took on average 22 minutes to complete. The survey included background factors such as demographics and life circumstances, a series of cognitive tasks, as well as self-report measures of time use, and motivations for video game play. In this paper, we focus on the playtime data and the self-reported well-being measures, and detail only those measures below. Other measures were intended for analyses beyond the scope of the current paper. All study data and materials are available on the OSF (see supplementary materials).

@@

```{r}
#| label: fig-playtime
#| fig-cap: "Description of playtime in the sample. Panel A shows the distribution of playtime across players for each observation period (time window leading up to the survey). Means are shown in red; hours of play for observation periods below the dashed line indicate total hours, whereas hours of play for periods above the dashed line reflect hours of play per day. Panel B shows the proportion of players who logged at least one play session in that period. Raw descriptive statistics are available in the supplementary materials (tables-figures/playtimeDescriptives.csv)."
#| fig-width: 8
#| fig-height: 4
#| fig-align: center
#| fig-pos: "t"
#| apa-twocolumn: true

@@

However, we did find evidence of direct positive relationships linking life fit to well-being separate from playtime (@fig-lifefit): Those who believe gaming to be beneficial to their lives tend to are also more likely to report higher levels of well-being, regardless of how much they play. Across 48 models, we observed relationships between well-being and a 1 point change in life fit ranging from `r min(directEsts$estimate)` to `r max(directEsts$estimate)` (median = `r round(median(directEsts$estimate), 3)`; all ps \< .001). In sensitivity analyses, life fit was both a significant predictor of daily life need satisfaction (`r nsMod$pretty`) and frustration (`r nfMod$pretty`), and largely remained a significant predictor of well-being when adjusting for need satisfaction and frustration (median b = `r round(median(fittedNSModels$estimate), 3)`, median p value = `r round(median(fittedNSModels$p.value), 3)`). This suggests that gaming life fit may account for well-being above and beyond someone's level of fulfillment in life as a whole.

@@

Although we did not intend our study to test the causal question most critical to heated debates surrounding video game engagement and global health, our study is a concrete step towards long-repeated calls for greater transparency and robustness in games research, having independently recruited a large sample of video game players (as opposed relying on recruitment through games companies themselves; e.g. @JohannesEtAl2021video), collected validated measures of well-being, joined these with digital trace data and made minimal adjustments for age, gender, employment, and education. Using these methods, we did not uncover robust or consistent relationships between time spent playing and various mental well-being outcomes.

@@

To further elucidate this point, we conducted brief simulation tests to ascertain how strong such confounding might need to be (see Supplementary Materials). For example, if the true standardized effect of playtime on mental health was a moderate .2 SDs per additional hour of daily playtime, a confound C would need to be a very strong cause of both X ($\beta = .5$) and Y ($\beta = -.5$ ) to bias the true .2 effect to null. While we do not claim this is impossible, we do believe it unlikely. Approaching the topic along these lines---identifying confounds, testing the presence or absence of correlations for their sensitivity to potential confounds, and systematically identifying factors that do (not) cause playtime and well-being---can help us achieve more systematic progress [@BallouEtAl2024How].This work can be bolstered by qualitative research aimed at more fully mapping the causal system and by substantive theory development with greater specificity in the aspects of media use expected to produce effects, the hypothesized causal relationships, boundary conditions, and so forth [@MagnussonEtAl2024Harmful; @Ballou2023Manifesto; @Coenen2023Lost; @EronenBringmann2021theory].

@@

As the field progresses, however, differences in the level of engagement pose major challenges for study sampling and generalizability. Calls for representative samples need to specify the population of reference: should this be the general population (of whom many do not play games), people who play any games at all (of whom many do not play the games for which researchers have data access), people who play the particular game or platform of interest (of whom many may be only minimally engaged), or something else entirely? In the field's quest for more generalizable results, this will be a critical issue.

@@

While we are quick to caution that the extremely sparse data at short timescales means that we cannot draw any meaningful conclusions, our data may be able to inform future hypotheses about how quickly the effects of playtime might materialize and fade---point estimates indicated that playtime was more strongly linked with greater well-being in the 1-2 hours prior to survey completion. This finding is compatible with various causal explanations: for example, players who recently played are more likely be in a period of leisure time, which would be expected to generate more positive feelings than in peers doing obligatory activities such as work. However, if researchers do expect to see positive effects of gaming, our data suggest that they may need to search for highly proximal effects directly during and following a play session [e.g., @VuorreEtAl2024Affective].

Should future studies find effects to disappate on similar timescales, it would go a long way towards explaining the juxtaposition of previous null findings from studies that related well-being to playtime over timescales such as two weeks [@VuorreEtAl2022Time], one month [@SibillaEtAl2021Harmonious], six months [@WeinsteinEtAl2017prospective], and one year [@KowertEtAl2015Psychosocial]---alongside other studies showing small relationships over the course of a day [@Allen2020gaming; @Razum2023Daily]. For most players, it may be the case that gaming is a recovery activity that helps to manage day-to-day stresses and mood fluctuations, without necessarily having substantial long-term impacts. The majority of players have several options for activities in their environment that would have comparable effects on their well-being. Such activities are thus “exchangeable”, serving the same short-term goals without consequences for long-term adjustment. Studying relationships over the course of hours has to date been possible largely only in laboratory settings—rarely have researchers had access to session-level data of naturalistic behavior that they could link to momentary well-being.

@@

There are four limitations and constraints on generalizability the merit mention. First, we could not analyse digital trace data generated when players engaged with third-party titles (i.e. games not published by Nintendo or its closely associated companies). As a result, our findings only hold for similar Nintendo games (75% of which were rated for everyone or everyone over 10), and it is not possible to rule out the possibility that third-party games with different content or themes might show a different pattern of effects.

@@

The idea that time spent playing is the key ingredient in how games impact well-being will be with us for some time. Although our study was designed to test an association rather than a causal link, it challenges the notion that simply playing more affects well-being, for better or for worse. The correlations we observed were mostly too small to practically matter. Moreover, we show that profound confounding would be required to suppress a true causal effect to account for the null associations we report. This is improbable but not impossible, and we believe our results agree with calls for scholars and health practitioners to embrace the gradual shift towards focusing on the quality, rather than quantity of video game play as the key factor for player health---in the first instance, this might be as simple as asking players how they feel gaming fits into and differentially affects other aspects of their life such as work/school, sleep, emotion regulation, and so on. This can be done in the context of self-report questionnaires, qualitative research, diagnostic interviews, and conversations between players and loved ones. To parlay this feedback from players into more systematic theory development, we encourage researchers to familiarize themselves with the breadth of proposed mechanisms through which different patterns—rather than hours—of gaming can affect people [e.g., @BallouEtAl2024How].
